### 2017-07-03 reasoning.tex
Soundness result added.

Two rules of the transaction-local RG judgment are shown in
Fig.~\ref{fig:rg-rules}. The rule \rulelabel{RG-Update} is
illustrative of the RG rules for SQL statements; they basically
reflect the structure of the corresponding reduction rule from
Fig.~\ref{fig:txnimp}, and require no further explanation. The rule
\rulelabel{RG-Foreach} defines the RG judgment for a \C{FOREACH} loop.
As is characteristic of loops, the reasoning is pivoted on a loop
invariant $\I$ (not same as the high-level invariant), that needs to
be stable w.r.t $\R$. $I$ must follow $P$, the pre-condition of
\C{FOREACH}, when no elements have been iterated, i.e, when
$y=\emptyset$. The body of the loop can assume the loop invariant, and
the fact that $z$ is an element from the set $x$ being iterated, to
prove its post-condition $Q_c$. Operational semantics ensures that $z$
is added to $y$ at the end of the iteration, hence $Q_c \conj z\in y$
is valid at the end of the iteration, from which $I$ must follow. When
\C{FOREACH} has finished execution, $y$, the set of iterated items, is
the entire set $x$. Thus $I \conj y=x$ must imply the post-condition
$Q$, which also needs to be stable. The rules for conditionals,
sequencing etc., are more-or-less standard, hence elided.

@@

@@

% \begin{definition}[\bfseries Interleaved step and multi-step relations]
% An interleaved step relation interleaves transaction-local reductions with
% legal interference from concurrent transactions as defined by the
% $\I$-constrained rely relation ($\R$).  It is defined thus:
% \begin{smathpar}
% \begin{array}{lcl}
% (c,\stl,\stg) \rstepsto (c',\stl',\stg') & \defeq & \stg \vdash 
%   (c,\stl) \stepsto (c',\stl') \conj \stg'=\stg \\
%   &   & \disj (c' = c \conj \stl'=\stl \conj \R(\stl, \stg, \stg'))\\
% \end{array}
% \end{smathpar}

\begin{definition}[\bfseries Interleaved step and multi-step relations]
An interleaved step relation interleaves a transaction's reduction with
interference from concurrent transactions as captured by the rely
relation. It is defined thus:
\begin{smathpar}
\begin{array}{lcl}
(c,\stg) \rstepsto (c',\stg') & \defeq &  
  (c,\stg) \stepsto (c',\stg') \disj (c' = c \conj R(\stg, \stg'))\\
\end{array}
\end{smathpar}

\noindent The interleaved step relation for transaction bound expressions
($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
reflexive transitive closure of the interleaved step relation.
\end{definition}

\begin{definition}[\bfseries Semantics of the RG judgment]
\label{def:rg-semantics}
The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
in terms of the interleaved step relation thus:\vspace*{-10pt}

\begin{smathpar}
\begin{array}{l}
\hspace*{-0.3in}
\rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
  \wedge \mathbb{I}(\E) \\
\hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
    \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
\hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
\end{array}
\end{smathpar}

\noindent The first conjunct in the consequent is called the \emph{Hoare
consequent} since it ascribes Hoare triple semantics to an RG sextuple.
The second conjunct, called the \emph{guarantee consequent}, uses the
$\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}

\begin{smathpar}
\begin{array}{l}
\texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
\hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
  (c'',\E'') \Rightarrow G(\E',\E'')\\
\end{array}
\end{smathpar}

\noindent The guarantee consequent requires $G$ to capture the trace effect of
every small-step of $c$, where the reduction can be interleaved by the
interference ($R$) from concurrent threads. The semantics of the RG
sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
similarly. Expressions, unlike commands, evaluate to a value $v$, and
the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
\end{definition}

Note that the semantics of all RG judgments, including the judgments
for transaction-bound terms, make similar demands of the guarantee
relation. Given that transactions are atomic (though not isolated), it
is not immediately apparent why a transaction's guarantee is required
to make explicit every step of its reduction. This requirement is
justified however because, in reality, a transaction's atomicity is
predicated on the isolation settings of the observer. A \iso{Read
  Uncommitted} transaction, for example, is permitted to observe the
internal state of a transaction $T$ even if $T$ is claimed to execute
atomically.  In the interest of modular verification, the transaction
must therefore make its internal state available via its guarantee
relation.

\begin{theorem}[\bfseries Soundness] 
The rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
is provided in the supplementary material.}
\end{theorem}

\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b) the effect
that every small-step of $c$ has on the trace is contained in $G$.
Soundness of the RG judgment for transaction-bound commands
($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
is always evaluated to a value that satisfies $C$.

>>>>

Two rules of the transaction-local RG judgment are shown in
Fig.~\ref{fig:rg-rules}. The rule \rulelabel{RG-Update} is
illustrative of the RG rules for SQL statements; they basically
reflect the structure of the corresponding reduction rule from
Fig.~\ref{fig:txnimp}, and require no further explanation. The rule
\rulelabel{RG-Foreach} defines the RG judgment for a \C{FOREACH} loop.
As is characteristic of loops, the reasoning is pivoted on a loop
invariant $I$ (not same as the high-level invariant), that needs to
be stable w.r.t $\R$. $I$ must follow $P$, the pre-condition of
\C{FOREACH}, when no elements have been iterated, i.e, when
$y=\emptyset$. The body of the loop can assume the loop invariant, and
the fact that $z$ is an element from the set $x$ being iterated, to
prove its post-condition $Q_c$. Operational semantics ensures that $z$
is added to $y$ at the end of the iteration, hence $Q_c \conj z\in y$
is valid at the end of the iteration, from which $I$ must follow. When
\C{FOREACH} has finished execution, $y$, the set of iterated items, is
the entire set $x$. Thus $I \conj y=x$ must imply the post-condition
$Q$, which also needs to be stable. The rules for conditionals,
sequencing etc., are more-or-less standard, hence elided.

@@

We now formalize the semantics of the RG judgments defined in
Fig.~\ref{fig:rg-rules}, and state their soundness guarantees.

@@

\begin{definition}[\bfseries Interleaved step and multi-step relations]
Interleaved step relations interleave global and transaction-local
reductions with interference as captured by the corresponding rely
relations. They are defined thus:
\begin{smathpar}
\begin{array}{lclr}
(c,\stg) \rstepsto (c',\stg') & \defeq &  
  (c,\stg) \stepsto (c',\stg') \disj (c' = c \conj R(\stg, \stg'))&
  \texttt{global}\\
(c,\stl,\stg) \rstepsto (c',\stl',\stg') & \defeq & \stg \vdash 
  (c,\stl) \stepsto (c',\stl') \conj \stg'=\stg& \texttt{transaction-local}\\
  &   & \disj (c' = c \conj \stl'=\stl \conj \R(\stl, \stg, \stg'))
\end{array}
\end{smathpar}

\noindent An interleaved multi-step relation ($\stepssto{n}$) is the
step-indexed reflexive transitive closure of the interleaved step
relation.  
\end{definition}

\begin{definition}[\bfseries Semantics of RG judgments]
\label{def:rg-semantics}
The semantics of the global and transaction-local RG judgments are
defined thus:
\begin{smathpar}
\begin{array}{lcl}
\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg'.~
  P(\stl,\stg) \conj (c,\stl,\stg) \rstepssto{n} (\cskip, \stl',\stg')
  \Rightarrow Q(\stl',\stg')\\
\rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
  \Rightarrow (\forall n,\stg'.\; (c,\stg) 
    \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
&&\hspace*{0.5in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
\end{array}
\end{smathpar}

\noindent The
$\texttt{step-guaranteed}$ predicate used in the semantics of the
global RG judgment is defined below:\vspace*{-10pt}

\begin{smathpar}
\begin{array}{lcl}
\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall n,\stg',c'',\stg''.
(c,\stg) \rstepssto{n} (c',\stg') \conj (c',\stg') \stepsto
  (c'',\stg'') \Rightarrow G(\stg',\stg'')\\
\end{array}
\end{smathpar}
\end{definition}

We can now assert the soundness of the RG judgments in
Fig.~\ref{fig:rg-rules} as following\footnote{Formal proof of soundness
is provided in the supplementary material.}:

\begin{theorem}[\bfseries Soundness] 
The rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
Definition~\ref{def:rg-semantics}.
\end{theorem}

\noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)
every interleaved multi-step reduction of $c$ preserves the database
integrity constraint $I$, and (b) the effect that every transaction in
$c$ has on the database state is captured by $G$.
